---
title: "Enhanced SCM with Interventions for Margotsphere"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Enhanced SCM with Interventions for Margotsphere}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The enhanced SCM functionality in `margot.core` provides a comprehensive framework for specifying structural causal models with:

1. **Mandatory exogenous errors** - Every endogenous node must have an explicit exogenous term
2. **Timeline DSL** - User-friendly domain-specific language for model specification
3. **Built-in shift interventions** - Store intervention policies directly in the SCM object

All functions follow the `margot_` naming convention for consistency across the margotsphere ecosystem.

```{r setup}
library(margot.core)
```

## Timeline DSL: User-Friendly Model Specification

The timeline DSL provides multiple intuitive ways to specify temporal causal models:

### Method 1: Simple names with K parameter (NEW!)

```{r dsl-simple}
# Simple variable names with K parameter
dsl <- "
U = U_l, U_a, U_y
l ~ U_l
a ~ l + U_a
y ~ a + l + U_y
"

scm <- margot_scm_from_dsl(dsl, K = 3)
print(scm)
```

### Method 2: Automatic detection from variable names

```{r dsl-auto}
# Variables with t{n}_ prefix auto-detect their time
dsl <- "
U = U_l, U_a, U_y
t0_l ~ U_l
t1_a ~ t0_l + U_a
t2_y ~ t1_a + t0_l + U_y
"

scm <- margot_scm_from_dsl(dsl)
print(scm)
```

### Method 3: Explicit k specification (original format)

```{r dsl-explicit}
# Explicit k specification for full control
dsl <- "
U  = U_l, U_a, U_y
k1 = t0_l ~ U_l
k2 = t1_a ~ t0_l + U_a
k3 = t2_y ~ t1_a + t0_l + U_y
"

scm <- margot_scm_from_dsl(dsl)
print(scm)
```

### DSL Rules

1. **One `U =` line** declaring all exogenous variables
2. **Flexible time specification**:
   - Use simple names with `K` parameter for automatic assignment
   - Use `t{n}_` prefix for automatic detection (t0 → k=1, t1 → k=2, etc.)
   - Use explicit `k# =` for full control
3. **Automatic exogenous mapping** - parser matches U terms from parents

## Shift Interventions

Store multiple intervention policies directly in the SCM:

```{r shifts}
# Add static shift
scm <- margot_scm_add_shift(
  scm,
  node = "t1_a",
  name = "always_treat",
  fun = function(a, ...) 1
)

# Add threshold intervention
scm <- margot_scm_add_shift(
  scm,
  node = "t1_a",
  name = "threshold_0.5",
  fun = function(a, ...) ifelse(a < 0.5, 0.5, a)
)

# Add dynamic shift based on covariates
scm <- margot_scm_add_shift(
  scm,
  node = "t1_a",
  name = "adaptive",
  fun = function(a, data, ...) {
    # increase treatment for high-risk (high L values)
    l_val <- data[["t0_l"]]
    ifelse(l_val > median(l_val), pmin(a + 0.2, 1), a)
  }
)

print(scm)
```

## Automatic SCM from Data

For existing data following LMTP conventions:

```{r from-data}
# Create example data
set.seed(2025)
n <- 100
df <- data.frame(
  t0_b1 = rnorm(n),
  t0_b2 = rbinom(n, 1, 0.4),
  t1_l = rnorm(n),
  t1_a = rbinom(n, 1, 0.5),
  t2_y = rnorm(n)
)

# Automatic SCM creation
scm_auto <- margot_scm_from_df(df)
print(scm_auto)
```

## Complex Models with Feedback

The DSL supports complex temporal dependencies:

```{r feedback}
feedback_dsl <- "
U = U_l, U_a, U_y
k1 = t0_l ~ U_l
k1 = t0_a ~ t0_l + U_a
k1 = t0_y ~ t0_a + t0_l + U_y
k2 = t1_l ~ t0_y + t0_a + U_l         # outcome affects future confounder
k2 = t1_a ~ t1_l + t0_y + U_a         # outcome affects future treatment
k2 = t1_y ~ t1_a + t1_l + t0_y + U_y
"

scm_feedback <- margot_scm_from_dsl(feedback_dsl)
print(scm_feedback)
```

## Integration with LMTP

Extract shift functions for use with `lmtp` package:

```{r lmtp-integration, eval=FALSE}
# Get shift function
shifts <- margot_scm_get_shifts(scm, "t1_a")
adaptive_shift <- shifts$adaptive

# Use with lmtp (assuming data is prepared)
# library(lmtp)
# result <- lmtp_tmle(
#   data = your_data,
#   trt = "t1_a",
#   outcome = "t2_y",
#   shift = adaptive_shift,
#   ...
# )
```

## LaTeX Output for Manuscripts

Generate publication-ready equations:

```{r latex}
cat(margot_scm_as_latex(scm))
```

## Reporting Function

A convenience function for clean SCM reports:

```{r reporting}
margot_report_scm <- function(scm, title = NULL) {
  if (!is.null(title)) {
    cat(title, "\n")
    cat(paste(rep("=", nchar(title)), collapse = ""), "\n\n")
  }
  
  cat("Structure:\n")
  cat("- Nodes:", length(scm$nodes), "\n")
  cat("- Time periods:", length(unique(scm$k)), "\n")
  
  n_interventions <- sum(sapply(scm$shift, length))
  if (n_interventions > 0) {
    cat("- Interventions:", n_interventions, "\n")
    for (node in names(scm$shift)) {
      cat("  *", node, ":", paste(names(scm$shift[[node]]), collapse = ", "), "\n")
    }
  }
  
  cat("\n")
  invisible(margot_scm_as_latex(scm))
}

# Use it
margot_report_scm(scm, "Analysis SCM")
```

## Extracting Components

Access specific parts of the SCM:

```{r extract}
# Get time indices
time_indices <- margot_scm_time_indices(scm)
print(time_indices)

# Get all shifts for a node
t1a_shifts <- margot_scm_get_shifts(scm, "t1_a")
cat("\nShifts available for t1_a:\n")
str(t1a_shifts, max.level = 1)
```

## Best Practices

1. **Always specify exogenous terms** - The parser enforces this for identification
2. **Use meaningful shift names** - Makes analysis code self-documenting
3. **Test shifts before adding** - Ensure they behave as expected
4. **Document complex dependencies** - Use comments in the DSL

## Advanced Example: Multi-Outcome Model

```{r multi-outcome}
multi_dsl <- "
U = U_b, U_l, U_a, U_y1, U_y2, U_m
k1 = b ~ U_b
k2 = l ~ b + U_l
k2 = a ~ b + l + U_a
k3 = m ~ a + l + U_m                  # mediator
k3 = y1 ~ a + l + m + U_y1            # outcome 1
k3 = y2 ~ a + l + U_y2                # outcome 2 (no mediator)
"

scm_multi <- margot_scm_from_dsl(multi_dsl)

# Add different shifts for different estimands
scm_multi <- margot_scm_add_shift(
  scm_multi, "a", "total_effect",
  function(a, ...) a + 0.5
)

scm_multi <- margot_scm_add_shift(
  scm_multi, "m", "direct_effect", 
  function(m, ...) m  # hold mediator constant
)

print(scm_multi)
```

This enhanced SCM framework provides a complete solution for specifying, manipulating, and analyzing structural causal models within the margotsphere ecosystem.