---
title: "Structural Causal Models in LMTP Style"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Structural Causal Models in LMTP Style}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `margot.core` package provides flexible helpers for creating and printing Structural Causal Models (SCMs) in the LMTP style (Díaz et al.). This functionality is designed to work seamlessly across the margotsphere ecosystem.

```{r setup}
library(margot.core)
```

## Creating SCMs from Data

The most common use case is creating an SCM from a wide data frame that follows the LMTP naming convention (`t0_*`, `t1_*`, ...).

```{r from-df}
# create example data
set.seed(2025)
df <- data.frame(
  # baseline
  t0_b1 = rnorm(100),
  t0_b2 = rbinom(100, 1, 0.3),
  t0_l = rnorm(100),
  
  # time 1
  t1_l = rnorm(100),
  t1_a = rbinom(100, 1, 0.5),
  t1_y = rnorm(100),
  
  # time 2
  t2_l = rnorm(100),
  t2_a = rbinom(100, 1, 0.5),
  t2_y = rnorm(100)
)

# create scm
scm <- scm_from_df(df)
print(scm)
```

## Creating SCMs from Equations

You can also specify an SCM explicitly using equation notation:

```{r from-spec}
spec <- "t0_b ~ U1 + U2; t1_a ~ t0_b + U1; t1_y ~ t1_a + U2"
scm2 <- scm_from_spec(spec)
print(scm2)
```

## LaTeX Output

For manuscripts and presentations, you can generate LaTeX representations:

```{r latex}
cat(as_latex(scm2))
```

This produces LaTeX code that can be included in your documents:

```latex
\begin{align*}
t0_b &:= f_{t0_b}(\,U1, U2, u_{t0_b}) \\
t1_a &:= f_{t1_a}(\,t0_b, U1, u_{t1_a}) \\
t1_y &:= f_{t1_y}(\,t1_a, U2, u_{t1_y}) \\
\end{align*}
```

## Working with Complex Models

### Time-Varying Confounding with Feedback

```{r feedback}
# model with treatment-confounder feedback
feedback_spec <- paste(
  "t0_l ~ U_l;",
  "t0_a ~ t0_l + U_a;",
  "t0_y ~ t0_a + t0_l + U_y;",
  "t1_l ~ t0_y + t0_a + U_l;",     # y affects future l
  "t1_a ~ t1_l + t0_y + U_a;",     # y affects future a
  "t1_y ~ t1_a + t1_l + t0_y + U_y",
  sep = " "
)

scm_feedback <- scm_from_spec(feedback_spec)
summary(scm_feedback)
```

### M-Bias Structure

```{r mbias}
# classic m-bias structure
m_bias <- scm_from_spec("U1 ~ ; U2 ~ ; X ~ U1; Y ~ U2; M ~ U1 + U2")
print(m_bias)
```

## Integration with LMTP Workflows

The SCM objects are designed to integrate with LMTP estimation workflows:

```{r integration, eval=FALSE}
# extract parent sets for LMTP
scm <- scm_from_df(your_data)

# use parent sets to define history for LMTP
# for node "t2_a", its parents define the adjustment set
adjustment_vars <- scm$pa[["t2_a"]]

# use in lmtp call
# lmtp::lmtp_tmle(
#   data = your_data,
#   trt = "t2_a",
#   outcome = "t2_y",
#   baseline = adjustment_vars,
#   ...
# )
```

## Summary Statistics

Get insights into your causal structure:

```{r summary}
# create a more complex model
complex_df <- data.frame(
  t0_b1 = rnorm(10), t0_b2 = rnorm(10), t0_b3 = rnorm(10),
  t0_l = rnorm(10), t0_a = rbinom(10, 1, 0.5),
  t1_l = rnorm(10), t1_a = rbinom(10, 1, 0.5), 
  t1_y1 = rnorm(10), t1_y2 = rnorm(10),
  t2_l = rnorm(10), t2_a = rbinom(10, 1, 0.5),
  t2_y1 = rnorm(10), t2_y2 = rnorm(10)
)

scm_complex <- scm_from_df(complex_df)
summary(scm_complex)
```

## Customizing Parent Sets

While `scm_from_df()` uses sensible defaults based on LMTP conventions, you might want to customize the parent sets:

```{r customize}
# start with automatic structure
scm <- scm_from_df(df)

# modify parent sets
# remove some dependencies for t2_y
scm$pa$t2_y <- c("t1_a", "t2_a")  # only depends on treatments

# create new scm with modified structure
scm_modified <- new_scm(
  nodes = scm$nodes,
  parents = scm$pa,
  exogenous = scm$U
)

print(scm_modified)
```

## Best Practices

1. **Naming Convention**: Follow the LMTP naming convention (`t{k}_{block}`) for automatic SCM generation
2. **Validation**: Use `summary()` to check the dependency structure makes sense
3. **Documentation**: Generate LaTeX output for papers to clearly communicate your causal assumptions
4. **Integration**: Use the parent sets from SCM objects to define adjustment sets in your analyses